---
title: 'Week 8 class: Data types'
output:
  github_document
date: "2024-03-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data types in R

### Vectors


#### Numeric vectors

Numeric vectors contain numbers.

```{r}
x <- 1:5

y <- c(5, 17, 87, 145, -9, 0)

x * 2

x + y

```

If you want to produce pdf files directly in R Markdown either install Latex or the **tinytex** package.


Missing values in R are denoted as NA.

```{r}
z <- c(1:5, NA, 6:10)
z

w <- c(1:5, NaN, 6:10)
w

# When doing the analysis sometimes you need to explicitly exclude the missing values.

mean(z)
mean(z, na.rm = TRUE)

```

#### Logical vectors

```{r}
# T is for True
# F is for False

x <- c(TRUE, TRUE, FALSE, FALSE)
x

z > 3

TRUE == 1
FALSE == 0

mean(x)

```

#### Character vectors

```{r}
a <- c("James", "aads", "Julia")
a

a <- c("James", "aads", "Julia", NA)
a

a <- c("James", "aads", "Julia", NA, 1)
a

```

#### Checking type of the object

```{r}
class(z)
typeof(z)
str(z)

class(x)
class(a)

```

#### Changing the type of vectors

```{r}
x <- c("1", "2", "3", "a")
class(x)
as.numeric(x)
x <- as.numeric(x)
class(x)
x

as.character(x)

# Number of elements in a vector

length(a)

```

### Data frames

```{r}
df <- data.frame(name1 = 1:5, name2 = rep(TRUE, 5), name3 = letters[1:5])
df

colnames(df)
rownames(df)

dim(df)

class(df)

# subsetting
df[1, 1]
df[2, 1]
df[1, 2]


# tibble
library(tidyverse)
df <- tibble(df)
class(df)

# accessing columns in a data frame
# name1: doesn't work
df$name1

```

### Matrices

```{r}
m <- matrix(1:9, nrow = 3)
m

m[1:2, 1:2]

class(m)

```

### Lists

```{r}
l1 <- list(x, y, z, w, a, df, m)
l1  

class(l1)
typeof(l1)


class(l1[1])
class(l1[[1]])

  
```

### Factors

```{r}
sex <- c(1, 2, 2, 2, 1, 1, 2, 1)
class(sex)

# recode into character vector
sex_ch <- if_else(sex == 1, "man", "woman")
class(sex_ch)

sex_fc <- factor(sex_ch, levels = c("man", "woman"))
sex_fc
class(sex_fc)

```

There are also dates and times, and that's a separate data type -- see the textbook.

Exercise:

Open the m_indresp data file and regress life satisfaction on sex and age. Interpret the results. Extract the coefficients from the model object and save as a vector. Extract the R-squared.

```{r}
library(vroom)
m_indresp <- vroom("UKDA-6614-tab/tab/ukhls/m_indresp.tab")
head(m_indresp)

m_indresp <- m_indresp |>
  select(pidp, m_sex_dv, m_age_dv, m_sclfsato)

m_indresp |>
  count(m_sex_dv)

m_indresp |>
  count(m_age_dv) |>
  print(n = Inf)

m_indresp |>
  count(m_sclfsato) |>
  print(n = Inf)

m_indresp <- m_indresp |>
  mutate(m_sclfsato = if_else(m_sclfsato > 0, m_sclfsato, NA_real_)) |>
  mutate(m_age_dv = if_else(m_age_dv > 0, m_age_dv, NA_real_)) |>
  mutate(m_sex_dv = if_else(m_sex_dv > 0, m_sex_dv, NA_real_)) 

# or:
# m_indresp <- m_indresp |>
#    mutate(across(c(m_sclfsato, m_age_dv, m_sex_dv), ~ if_else(.x > 0, .x, NA_real_)))

model1 <- lm(m_sclfsato ~ m_sex_dv + m_age_dv + I(m_age_dv ^ 2), data = m_indresp)

model1

class(model1)
typeof(model1)

str(model1)

# Extract the coefficients
model1$coefficients
model1[1]
model1[[1]]

# Extract R-squared

str(summary(model1))
summary(model1)$r.squared
summary(model1)[[8]]

# m_indresp |>
#   mutate(across(m_sex_dv:m_sclfsato, if_else(. > 0, ., NA_real_)))
#     
    

```

2. Recode age into age groups. Convert age into a factor, with "16-35" as the reference category. Regress life satisfaction on sex and age groups again. What changed in the interpretation.

```{r}
m_indresp <- m_indresp |>
  mutate(m_age_gr = case_when(
    m_age_dv %in% 16:34 ~ "16-34",
    m_age_dv %in% 35:59 ~ "35-59",
    m_age_dv >= 60  ~ ">=60"
  )) |>
  mutate(m_age_gr = factor(m_age_gr, levels = c(">=60", "16-34", "35-59")))

model2 <- lm(m_sclfsato ~ m_sex_dv + m_age_gr, data = m_indresp)
summary(model2)

m_indresp  <- m_indresp |>
  mutate(m_age_gr = fct_relevel(m_age_gr, "16-34", "35-59"))

lm(m_sclfsato ~ m_sex_dv + m_age_gr, data = m_indresp) |>
  summary()

```

How life satisfaction depends on age

```{r}
m_indresp |>
  ggplot(aes(x = m_age_dv, y = m_sclfsato)) +
  geom_smooth()

```


For week 9: read the Data import chapter from the R for Data Science textbook: https://r4ds.hadley.nz/import